15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 2

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling and successional trajectories [@lovett_nonnative_2016; @herms_emerald_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

@@

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context. EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list). The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022]. Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threats_2020].

>>>>

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling and successional trajectories [@lovett_nonnative_2016; @herms_emerald_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013a; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion by other nuisance species [@fei_biomass_2019].

@@

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context. EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list). The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022]. Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].
